Notes on Perturbational Quantities

### 2011-06-15 docs/pert_notes/pert.tex
start reworking this to allow for other species_pred_types

%-----------------------------------------------------------------------------
% density
%-----------------------------------------------------------------------------
\section{Density Evolution}
The full density evolution equation is
\begin{eqnarray}
\frac{\partial\rho}{\partial t} &=& -\nabla\cdot(\rho\Ub) \nonumber \\
&=& -\Ub\cdot\nabla\rho - \rho\nabla\cdot\Ub. \label{rho equation}
\end{eqnarray}
We predict both $\rho'$ and $\rho_0$ to edges separately and later use them to 
reconstruct $\rho$ at edges.  The base state density evolution equation is
\begin{equation}
\frac{\partial\rho_0}{\partial t} = -\nabla\cdot(\rho_0 w_0 \eb_r) = 
-w_0\frac{\partial\rho_0}{\partial r} 
\underbrace{-\rho_0\frac{\partial w_0}{\partial r}}_{"\rho_0 ~ \text{force}"}.
\label{rho0 equation}
\end{equation}
Subtract (\ref{rho0 equation}) from (\ref{rho equation}) and rearrange
terms, noting that $\Ub = \Ubt + w_o\eb_r$, to obtain the
perturbational density equation,
\begin{equation}
\frac{\partial\rho'}{\partial t} = -\Ub\cdot\nabla\rho' \underbrace{- \rho'\nabla\cdot\Ub 
- \nabla\cdot(\rho_0\Ubt)}_{\rho' ~ \text{force}},
\label{rhoprime equation}
\end{equation}
where the underbraced "$\rho'$ force'' is computed in {\tt modify\_scal\_force}.

\subsection{Predicting $\rho'$ at edges}
We define $\rho' = \rho^{(1)} - \rho_0^n$.  Then we predict $\rho'$ to
edges using {\tt make\_edge\_scal} in {\tt scalar\_advance} and the
underbraced term in (\ref{rhoprime equation}) as the forcing.

\subsection{Predicting $\rho_0$ at edges}\label{Predicting rho0 at edges}
There are two ways to predict $\rho_0$ at edges.
\begin{enumerate}

@@

\subsection{Computing $\rho$ at edges}\label{Computing rho at edges}
For the non-radial edges, we directly add $\rho_0^{n+\myhalf,{\rm avg}}$
to $\rho'$ since $\rho_0^{n+\myhalf,{\rm avg}}$ is a cell-centered
quantity.  For the radial edges, we interpolate to obtain
$\rho_0^{n+\myhalf,{\rm avg}}$ at radial edges before adding it to
$\rho'$.

\subsection{Advancing $\rho X_k$}\label{Advancing rhoX_k}
The evolution equation for $\rho X_k$, ignoring the reaction terms
that were already accounted for in {\tt react\_state}, and the
associated discretization is
\begin{equation}
\frac{\partial\rho X_k}{\partial t} = -\nabla\cdot(\rho X_k\Ub) = 
-\nabla\cdot\left\{\left[\left(\rho_0^{n+\myhalf,{\rm avg}} 
+ \rho'\right)X_k\right](\Ubt+w_0\eb_r)\right\}.
\end{equation}

@@

\subsection{Predicting $(\rho h)'$ at edges}\label{Predicting rhohprime at edges}
We define $(\rho h)' = (\rho h)^{(1)} - (\rho h)_0^n$.  Then we predict 
$(\rho h)'$ to edges using {\tt make\_edge\_scal} in {\tt scalar\_advance} 
and the underbraced term in (\ref{rhohprime equation}) as the forcing.

>>>>

%-----------------------------------------------------------------------------
% density
%-----------------------------------------------------------------------------
\section{Density Evolution}
The full density evolution equation is
\begin{eqnarray}
\frac{\partial\rho}{\partial t} &=& -\nabla\cdot(\rho\Ub) \nonumber \\
&=& -\Ub\cdot\nabla\rho - \rho\nabla\cdot\Ub \, . \label{rho equation}
\end{eqnarray}
The species are evolved according to
\begin{eqnarray}
\frac{\partial(\rho X_k)}{\partial t} &=& -\nabla\cdot(\rho\Ub X_k) + \rho \omegadot_k \nonumber \\
&=& -\Ub\cdot\nabla\rho - \rho\nabla\cdot\Ub + \rho \omegadot_k \, . \label{species equation}
\end{eqnarray}
In practice, only the species evolution equation is evolved, and the
total density is set as
\begin{equation}
\rho = \sum_k (\rho X_k)
\end{equation}
To advance $(\rho X_k)$ to the next timestep, we need to predict a
time-centered, interface state.  Algebraically, there are multiple
paths to get this interface state---we can predict $(\rho X)$ to the
edges as a single quantity, or predict $\rho$ and $X$ separately
(either in full or perturbation form).  In the notes below, we use the
subscript `{\tt edge}' to indicate what quantity was predicted to the
edges.  In MAESTRO, the different methods of computing $(\rho X)$ on
edges are controlled by the {\tt species\_pred\_type} parameter.  The
resulting edge state is shown in the following table:

\begin{table}[h]
\centering
\begin{tabular}{c|c}
\hline
\hline
{\tt species\_pred\_type} &   {$(\rho X_k)$ edge state} \\
\hline \\[-3mm]
1 / {\tt pred\_rhoprime\_and\_X}  &  $\left(\rho_0^{n+\myhalf,{\rm avg}}
+ \rho'_\mathrm{edge} \right)(X_k)_\mathrm{edge}$ \\[2mm]
2 / {\tt pred\_rhoX}  &  $(\rho X_k)_\mathrm{edge}$ \\
\hline
\end{tabular}
\end{table}
We note the labels {\tt pred\_rhoprime\_and\_X} and {\tt pred\_rhoX} are
provided by the {\tt pred\_parameters} module.

\subsection{Advective force for interface prediction}
In predicting the interface states, we work with an advection-like
equation, and call anything other than the advective term a ``force''.
For example, for a scalar $\phi$, obeying a conservation law with a source term, $f$:
\begin{equation}
\frac{\partial \phi}{\partial t} = -\nabla \cdot (\phi \Ub) + f
\end{equation}
We write this in advective form as:
\begin{equation}
\frac{\partial \phi}{\partial t} = -\Ub \cdot \nabla \phi \, \underbrace{- \phi \nabla \cdot \Ub + f}_{\text{``advective force''}}
\end{equation}
Here we identify the non-advection portion of the divergence of the
flux together with the original force, $f$, as the ``force'' for the
advective equation.  In predicting the interface states, this advective
force will be used in computing the interface states.

\subsection{{\tt species\_pred\_type} = {\tt pred\_rhoprime\_and\_X}}

Here we wish to construct $(\rho_0^{n+\myhalf,{\rm avg}}        
+ \rho'_\mathrm{edge})(X_k)_\mathrm{edge}$.

We predict both $\rho'$ and $\rho_0$ to edges separately and later use them to 
reconstruct $\rho$ at edges.  The base state density evolution equation is
\begin{equation}
\frac{\partial\rho_0}{\partial t} = -\nabla\cdot(\rho_0 w_0 \eb_r) = 
-w_0\frac{\partial\rho_0}{\partial r} 
\underbrace{-\rho_0\frac{\partial w_0}{\partial r}}_{``\rho_0 ~ \text{force"}}.
\label{rho0 equation}
\end{equation}
Subtract (\ref{rho0 equation}) from (\ref{rho equation}) and rearrange
terms, noting that $\Ub = \Ubt + w_o\eb_r$, to obtain the
perturbational density equation,
\begin{equation}
\frac{\partial\rho'}{\partial t} = -\Ub\cdot\nabla\rho' \underbrace{- \rho'\nabla\cdot\Ub 
- \nabla\cdot(\rho_0\Ubt)}_{\rho' ~ \text{force}},
\label{rhoprime equation}
\end{equation}
where the underbraced "$\rho'$ force'' is computed in {\tt modify\_scal\_force}.

\subsubsection{Predicting $\rho'$ at edges}
We define $\rho' = \rho^{(1)} - \rho_0^n$.  Then we predict $\rho'$ to
edges using {\tt make\_edge\_scal} in {\tt density\_advance} and the
underbraced term in (\ref{rhoprime equation}) as the forcing.

\subsubsection{Predicting $\rho_0$ at edges}\label{Predicting rho0 at edges}
There are two ways to predict $\rho_0$ at edges.
\begin{enumerate}

@@

We note that in essence these choices reflect a hyperbolic (1)
vs.\ elliptic (2) approach.  In MAESTRO, if we setup a problem with
$\rho = \rho_0$ initially, and enforce a constraint $\nabla \cdot
(\rho_0 \Ub) = 0$ (i.e.\ the anelastic constraint), then analytically,
we should never generate a $\rho'$.  To realize this behavior
numerically, we use $\rho_0^{n+\myhalf,{\rm avg}}$ in the prediction
of $(\rho X_k)$ on the edges to be consistent with the use of the
average of $\rho$ to the interfaces in the projection step at the end
of the algorithm.

\subsubsection{Computing $\rho$ at edges}\label{Computing rho at edges}
For the non-radial edges, we directly add $\rho_0^{n+\myhalf,{\rm avg}}$
to $\rho'$ since $\rho_0^{n+\myhalf,{\rm avg}}$ is a cell-centered
quantity.  For the radial edges, we interpolate to obtain
$\rho_0^{n+\myhalf,{\rm avg}}$ at radial edges before adding it to
$\rho'$.

\subsubsection{Predicting $X_k$ at edges}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{{\tt species\_pred\_type} = {\tt pred\_rhoX}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Advancing $\rho X_k$}\label{Advancing rhoX_k}
The evolution equation for $\rho X_k$, ignoring the reaction terms
that were already accounted for in {\tt react\_state}, and the
associated discretization is
\begin{equation}
\frac{\partial\rho X_k}{\partial t} = -\nabla\cdot(\rho X_k\Ub) = 
-\nabla\cdot\left\{\left[\left(\rho_0^{n+\myhalf,{\rm avg}} 
+ \rho'\right)X_k\right](\Ubt+w_0\eb_r)\right\}.
\end{equation}

@@

\subsection{Predicting $(\rho h)'$ at edges}\label{Predicting rhohprime at edges}
We define $(\rho h)' = (\rho h)^{(1)} - (\rho h)_0^n$.  Then we predict 
$(\rho h)'$ to edges using {\tt make\_edge\_scal} in {\tt enthalpy\_advance} 
and the underbraced term in (\ref{rhohprime equation}) as the forcing.
